Short paper : Automatic predicate testing in formal certification

### 2016-02-21 ProofsAndTests/paper-TAP/content/1-introduction.tex
save

One way to increase the confidence in the software we build is to formally prove its correctness using a proof assistant. These proofs assistants enable to write code, logical statements and proofs in the same language, and offer the guarantee that every proof will be automatically checked. The appearance of these languages, and the way to write programs in them, varies from one to another. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantee tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function that works for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to only use a ``weak" type, like $sort\ :\ \{T:Type\}\ \rightarrow\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect? First, the output has to be sorted. So first, we need to define this notion of being sorted, here as an inductive predicate :

\begin{lstlisting}
data isSorted : {T:Type} -> (Tord : Order T) 
                 -> (l:List T) -> Type where
    NilIsSorted : (Tord : Order T) -> isSorted Tord []
    SingletonIsSorted : (Tord : Order T) -> (x:T) -> isSorted Tord [x]
    ConsSorted : {Tord : Order T} -> (h1:T) -> (h2:T) -> (l:List T) 
                 -> (isSorted Tord (h2::t)) -> (h1 $\leq$ h2) 
                 -> (isSorted Tord (h1::(h2::t)))
\end{lstlisting}
The first and second constructor of this predicate say that $[\ ]$ and $[x]$ are sorted according to any order, and for any $x$. The third one says that a list of two or more elements is sorted if the head $h1$ is lower or equal than the next element $h2$, and if the list deprived from its head is also sorted. In order to express that the result produced by the function $sort$ is sorted, we can write and prove the following lemma:
$sort\_correct : \forall\ (T:Type)\ (l:List\ T),\ isSorted\ (sort\ l)$. The problem with this specification is that it doesn't says anything about the content of the output. The function $sort$ could just return the empty list $[\ ]$ all the time, it would still be possible to prove this correctness lemma. Here, the problem is that the function is underspecified, and it is therefore possible to write a senseless implementation, which is unfortunately provably correct. Only a careful reader could realize that the lemma $sort\_correct$ forgets to mention that the input and output list should be in bijection, meaning that everything which was originally in the input list should still be in the output, and that nothing else should have been added.

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Then, the function $sort$ could do whatever it wants, its result would be seen as ``sorted". Said differently, any list would be seen as ``sorted" according to this predicate, just because of this single typo, and the algorithm could for example return its input unchanged. One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is that currently, many effort are going in the direction of proof automation, where the machine builds the proof automatically. Many proof assistants are enhanced regularly with new automations that solves automatically some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures~\cite{Slama2016}. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? The primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show some basic approaches to the problem of underspecification (section~\ref{sect:naiveApproaches})
	\item Present a way to test in the proof assistant whether the predicate effectively captures the notion that we wanted to define, by automatically generating terms, and we automate these tests as far as possible (section~\ref{sect:testingInside})
	\item Present how we can go a step forward by replacing these tests about the predicate by some proofs (section~\ref{sect:aStepForward})
\end{itemize}

>>>>

One way to increase the confidence in the software we build is to formally prove its correctness using a proof assistant. Proofs assistants enable to write code, logical statements and proofs in the same language, and offer the guarantee that every proof will be automatically checked. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantees tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to define the sorting function with a ``weak" type, like $sort\ :\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect? First, the output has to be sorted, so we need to define this notion of being sorted, here as an inductive predicate :

\begin{lstlisting}
data isSorted : {T:Type} -> (Tord : Order T) 
                 -> (l:List T) -> Type where
    NilIsSorted : (Tord : Order T) -> isSorted Tord []
    SingletonIsSorted : (Tord : Order T) -> (x:T) -> isSorted Tord [x]
    ConsSorted : {Tord : Order T} -> (h1:T) -> (h2:T) -> (t:List T) 
                 -> (isSorted Tord (h2::t)) -> (h1 $\leq$ h2) 
                 -> (isSorted Tord (h1::(h2::t)))
\end{lstlisting}
The first and second constructor of this predicate say that $[\ ]$ and $[x]$ are sorted according to any order, and for any $x$. The third one says that a list of two or more elements is sorted if $h1 \le h2$, and if the list deprived from its head is also sorted. In order to express that the result produced by the function $sort$ is sorted, we can write and prove the following lemma:
$sort\_correct : \forall\ (T:Type)\ (Tord:Order\ T)\ (l:List\ T),\ isSorted\ Tord\ (sort\ l)$. The problem with this specification is that it doesn't says anything about the content of the output. The function $sort$ could just return the empty list $[\ ]$ all the time, it would still be possible to prove this correctness lemma. Here, the problem is that the function is underspecified, and it is therefore possible to write a senseless implementation, which is unfortunately provably correct. Only a careful reader could realise that the lemma $sort\_correct$ forgets to mention that the input and output list should be in bijection, meaning that everything which was originally in the input list should still be in the output, and that nothing else should have been added.

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. With this mistake, any list would be seen as ``sorted", just because of this single typo, and the algorithm could for example return its input unchanged. One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is quite different because many effort are going in the direction of proof automation, which aims to let the machine generate automatically the proof for some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures~\cite{Slama2016}. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? The primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show some basic approaches to the problem of underspecification (section~\ref{sect:naiveApproaches})
	\item Present a way to test the predicate in the proof assistant, by automatically generating terms, and we automate these tests as far as possible (section~\ref{sect:testingInside})
	\item Present how we can go a step forward by replacing these tests about the predicate by some proofs (section~\ref{sect:aStepForward})
\end{itemize}

### 2016-02-21 ProofsAndTests/paper-TAP/content/2-specCanGoWrong.tex
save

When confronted to this problem of adequacy between the intuitive notion and the formalised one, a first possibility that usually requires quite a lot of effort is to formalise the notion multiple times, with different predicates, and to prove that they are equivalent.
With our example, that means that we need to find another formalisation $isSorted'$ of being sorted, and to prove the following lemma.
$pred\_equiv : \forall\ (T:Type)\ (l:List\ T), (isSorted\ l) \leftrightarrow (isSorted'\ l)$.
This approach aims at increasing the confidence in our formal definitions by assuming that if we've managed to define multiple times the same notion, then we've surely succeeded to define the notion we wanted. The biggest problem with this approach is to be able to find some alternative formalisations that are sufficiently different from the original one. Obviously, if the new normalisations are too similar to the original one (and in the worst case the new ones are just syntactical variants of the first one), then we won't gain any guarantee. The ideal would be to capture the same notion by using very different point of view, and we will show in section~\ref{sect:testingInside} an original approach for doing so.

@@

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) -> Dec(isSorted Tord v)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          with (decideIsSorted Tord (h2::t))
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (Yes h2_tail_sorted) = Yes (ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2)
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (No h2_tail_not_sorted) = No [...]
  decideIsSorted Tord (h1::(h2::t)) | No h1_not_lower_h2 = No [...]
\end{lstlisting}
Now, in order to do \emph{tests by proof}, we can simply run the decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}
And if we ask to evaluate $isSorted\_test1'$, the system will answer $Yes$ and the automatically generated proof of $isSorted\ [3,\ 5,\ 7]$, which is what we want, and we therefore know that the predicate has passed this test.

With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted --according to our intention-- but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms.

>>>>

When confronted to this problem of adequacy between the intuitive notion and the formalised one, a first possibility is to formalise the notion multiple times, with different predicates, and to prove that they are equivalent.
With our example, that means that we need to find another formalisation $isSorted'$ of being sorted, and to prove the following lemma.
$pred\_equiv : \forall\ (T:Type)\ (l:List\ T), (isSorted\ l) \leftrightarrow (isSorted'\ l)$.
This approach aims at increasing the confidence in our formal definitions by assuming that if we've managed to define multiple times the same notion, then we've surely succeeded to define the notion we wanted. The biggest problem with this approach is to be able to find some alternative formalisations that are sufficiently different from the original one. Obviously, if the new normalisations are too similar to the original one (and in the worst case the new ones are just syntactical variants of the first one), then we won't gain any guarantee. The ideal would be to capture the same notion by using very different point of view, and we will show in section~\ref{sect:testingInside} an original approach for doing so.

@@

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) -> Dec(isSorted Tord l)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          with (decideIsSorted Tord (h2::t))
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (Yes h2_tail_sorted) = Yes (ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2)
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (No h2_tail_not_sorted) = No [...]
  decideIsSorted Tord (h1::(h2::t)) | No h1_not_lower_h2 = No [...]
\end{lstlisting}
Now, in order to do \emph{tests by proof}, we can simply run the decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}
And if we ask to evaluate $isSorted\_test1'$, the system will answer $Yes$ and the automatically generated proof of $isSorted\ [3,\ 5,\ 7]$, which is what we want, and we therefore know that the predicate has passed this test. With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms.

### 2016-02-21 ProofsAndTests/paper-TAP/content/3-testingInside.tex
save

It is often easier to generate examples of a notion than it is to precisely define it. We can operate an interesting change of point of view by generating the set of terms that we precisely wanted to describe with the predicate. With our example of sorted lists, that means that we need to define the same notion of being sorted, but this time with a definition example-based. Writing a function that produces all the sorted lists might be a bit more difficult than just giving a few examples, but this complicated function will have the main advantage of being so different from the predicate that if the two notions agree, then we will have gained a great confidence on the predicate. We decide to use coinduction and the type Stream (a coinductive version of lists, potentially infinite) in order to generate all the sorted lists of size $n$ with what we call a generator.

@@

To do so, the type $T$ needs to be $RecursivelyEnumerable$, which means that there should there should exists a computational map $Nat\ \rightarrow\ Maybe\ c$ with the condition that this map is surjective, which means that all value of type $c$ should be hit at least once by the map : $map\_is\_surjective : (y:c)\ \rightarrow\ (x:Nat\ **\ (computableMap\ x\ =\ Just\ y))$.

@@

\begin{lstlisting}
testSorted : (m:Nat) -> (n:Nat) -> Vect m Bool
testSorted m n = 
  let x = generateSortedList T TisRecEnu TisOrdered n in
    let y = LLmap (\l => let res = decideIsSorted TisOrdered l in
			    case res of
			      Yes _ => True
			      No _ => False) x in
	unfold_n_times_with_padding y m True
\end{lstlisting}
And we can inspect the result of running the first 8 tests by evaluating $testSorted\ 8\ 5$.

>>>>

The key idea that this paper wants to convey is that it is often easier to generate examples of a notion than it is to precisely define it. We can operate an interesting change of point of view by generating the set of terms that we precisely wanted to describe with the predicate. With our example of sorted lists, that means that we need to define the same notion of being sorted, but this time with a definition example-based. Writing a function that produces all the sorted lists might be a bit more difficult than just giving a few examples, but this complicated function will have the main advantage of being so different from the predicate that if the two notions agree, then we will have gained a great confidence on the predicate. We decide to use coinduction and the type Stream (a coinductive version of lists, potentially infinite) in order to generate --with what we call a generator-- all the sorted lists of size $n$.

@@

To do so, the type $T$ needs to be $RecursivelyEnumerable$, which means that there should exists a computational map $Nat\ \rightarrow\ Maybe\ c$ with the condition that this map is surjective, which means that all value of type $c$ should be hit at least once by the map : $map\_is\_surjective : (y:c)\ \rightarrow\ (x:Nat\ **\ (computableMap\ x\ =\ Just\ y))$.

@@

\begin{lstlisting}
testSorted : (m:Nat) -> (n:Nat) -> Vect m Bool
testSorted m n = 
  let x = generateSortedList T TisRecEnu TisOrdered n in
    let y = Smap (\l => let res = decideIsSorted TisOrdered l in
			    case res of
			      Yes _ => True
			      No _ => False) x in
	unfold_n_times_with_padding y m True
\end{lstlisting}
And we can inspect the result of running the first 8 tests by evaluating $testSorted\ 8\ 5$.

### 2016-02-21 ProofsAndTests/paper-TAP/content/4-differentKindOfProperties.tex
save

In order to go a step forward, we can decide to replace the tests on the predicate by proofs. Instead of testing the predicate on a finite subset of the generated sorted lists, as we did in the previous section, we can prove that any of the automatically generated sorted list is provably sorted according to $isSorted$.

@@

This lemma has the advantage of not requiring the predicate to be decidable, when this was needed in the previous section where we tested the predicate on a finite observation. However, one could object that this lemma is itself built by using a predicate, $All$, and that we can't necessary trust blindly such a specification. The answer is that no guarantee is perfect, and all we can do is to add more guarantee, but there is always something to trust. Moreover, this new kind of specifications and proofs --about the predicate itself-- use more primitive components like streams and the predicate $All$, and these components can be provided once and for all. If they are part of some standard library being used intensively, there is very low risk that they don't capture the desired semantic.

In the other direction, we can prove that when the predicate $isSorted$ holds on some list, then this list is one of the list produced by the generator. The proof is done by induction on the list. The case of an empty list is trivial. For a singleton $[x]$, the proof uses the fact that the $T$ is recursively enumerable, and more precisely that the provided map is surjective, which means that it has hit any value, and thus in particular $x$. When there are two or more elements, we know that the smaller list deprived from its head is sorted according to $isSorted$ by construction. By using the lemma recursively on this smaller list together with the fact that the head is lower than the next element and the fact that the map is surjective, we get what we need. \qed

>>>>

In order to go a step forward, we can decide to replace the tests on the predicate by proofs. Instead of testing the predicate on a finite subset of all the generated sorted lists, as we did in the previous section, we can prove that any of the automatically generated sorted list is provably sorted according to $isSorted$.

@@

This lemma has the advantage of not requiring the predicate to be decidable, when this was needed in the previous section where we automatically tested the predicate on a finite observation. However, one could object that this lemma is itself built by using a predicate, $All$, and that we can't necessary trust blindly such a specification. The answer is that no guarantee is perfect, and all we can do is to add more guarantee, but there is always something to trust. Moreover, this new kind of specifications and proofs --about the predicate itself-- use more primitive components like streams and the predicate $All$, and these components can be provided once and for all. If they are part of some standard library being used intensively, there is very low risk that they don't capture the desired semantic.

In the other direction, we can prove that when the predicate $isSorted$ holds on some list, then this list is one of the list produced by the generator. The proof is done by induction on the list. For a singleton $[x]$, the proof uses the fact that $T$ is recursively enumerable, which means that the map has hit any value, and thus in particular $x$. When there are two or more elements, we know that $isSorted$ holds on the smaller list deprived from its head by construction. We finish the proof by using the lemma recursively on this smaller list together with the fact that the head is lower than the next element and the fact that the map is surjective.\qed
